---
title: "Ride-hailing driver study"
author: "Tukey Tu"
date: "10/4/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# UNDERSTANDING THE IMPACT OF A PANDEMIC ON PUBLIC TRANSIT USE PATTERNS

## import libraries
```{r echo = TRUE}
library(MASS)
library(lme4)
library(ggplot2)
library(psych)
library(dplyr)
library(MRCE)
library(car)
library(tidyverse)
library(ggpubr)
library(compare)
library(tibble)
library(tidyr)
library(gridExtra)
library(MASS)
library(Rmisc)
library(mgcv)
```

## set working space & Read the file
```{r}
setwd("/Users/tuyuanjie/Documents/2021Summer/Driverdata")

WideData <- read.csv("widedata.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)

```

## Quality check
```{R}
## attention check
WideData <- filter(WideData, ATTENTION == "N4")

## hhincome >= individual income
WideData <- filter(WideData, HHINCOME >= INDVINCOME)
```

## Data wrangling
```{R}
## only choose respondents from the airport
WideData <- filter(WideData, SOURCE == 3)

## filter out respondents who do not want to give the number (in the survey  < 18)
WideData <- filter(WideData, AGE >= 18)

## filter out respondents who do not want to give the number (in the survey recorded as 10)
WideData <- filter(WideData, INDVINCOME < 10)

## Age ==> categorical
WideData <- WideData %>% mutate(AGE = ifelse(AGE >= 65, 2,
                                              ifelse(AGE >= 40, 1, 0)))

## Education ==> categorical
WideData <- WideData %>% mutate(EDU = ifelse(EDU == 1, 0,
                                             ifelse(EDU == 2, 1,
                                                ifelse(EDU %in% c(3,4,5,8), 2, 3))))

## Ridehailing company
WideData$COMP_Uber[is.na(WideData$COMP_Uber)] <- 0
WideData$COMP_Lyft[is.na(WideData$COMP_Lyft)] <- 0


## Income ==> dummy
WideData <- WideData %>% mutate(POVERTY = ifelse(HHINCOME < 9, 0,1))

## Employment status ==> categorical
WideData$EMPL <- WideData$EMPL - 1

## filter out non-binary gender since too few
WideData <- filter(WideData, GENDER != 2)

## filter out people who choose "prefer not to answer"
WideData <- filter(WideData, EMPL %in% c(0,1,2))

## filter out people who completed no trips last week
WideData <- filter(WideData, WEEK.TRIPS!=0)

## calculate trip rejection rate
WideData$rjrate <- WideData$WEEK.TRIPS.RJ/(WideData$WEEK.TRIPS + WideData$WEEK.TRIPS.RJ)*100

## make ID categorical
WideData$ID <- as.numeric(as.factor(WideData$USERID))

```

## Wide to long: RESP
```{R}
obs1 <- WideData[,c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME", "INDVINCOME", "SOURCE", "RESP.1", "RESP.1.SURGE", "RESP.1.WTTIME", "RESP.1.RLTIME", "RESP.1.LTRIP", "RESP.1.PR", "RESP.1.SHARED", "rjrate")]
names(obs1) <- c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME", "INDVINCOME", "SOURCE", "RESP", "DMS", "DMW", "RT", "LT", "PR", "MD", "rjrate")
obs1$obs <- 1

obs2 <- WideData[,c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME", "INDVINCOME", "SOURCE", "RESP.2", "RESP.2.SURGE", "RESP.2.WTTIME", "RESP.2.RLTIME", "RESP.2.LTRIP", "RESP.2.PR", "RESP.2.SHARED", "rjrate")]
names(obs2) <- c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME", "INDVINCOME", "SOURCE", "RESP", "DMS", "DMW", "RT", "LT", "PR", "MD", "rjrate")
obs2$obs <- 2

obs3 <- WideData[,c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME","INDVINCOME",  "SOURCE", "RESP.3", "RESP.3.SURGE", "RESP.3.WTTIME", "RESP.3.RLTIME", "RESP.3.LTRIP", "RESP.3.PR", "RESP.3.SHARED", "rjrate")]
names(obs3) <- c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME", "INDVINCOME", "SOURCE", "RESP", "DMS", "DMW", "RT", "LT", "PR", "MD", "rjrate")
obs3$obs <- 3

obs4 <- WideData[,c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME","INDVINCOME",  "SOURCE", "RESP.4", "RESP.4.SURGE", "RESP.4.WTTIME", "RESP.4.RLTIME", "RESP.4.LTRIP", "RESP.4.PR", "RESP.4.SHARED", "rjrate")]
names(obs4) <- c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME", "INDVINCOME", "SOURCE", "RESP", "DMS", "DMW", "RT", "LT", "PR", "MD", "rjrate")
obs4$obs <- 4

obs5 <- WideData[,c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME", "INDVINCOME", "SOURCE", "RESP.5", "RESP.5.SURGE", "RESP.5.WTTIME", "RESP.5.RLTIME", "RESP.5.LTRIP", "RESP.5.PR", "RESP.5.SHARED", "rjrate")]
names(obs5) <- c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME", "INDVINCOME", "SOURCE", "RESP", "DMS", "DMW", "RT", "LT", "PR", "MD", "rjrate")
obs5$obs <- 5

obs6 <- WideData[,c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME", "INDVINCOME", "SOURCE", "RESP.6", "RESP.6.SURGE", "RESP.6.WTTIME", "RESP.6.RLTIME", "RESP.6.LTRIP", "RESP.6.PR", "RESP.6.SHARED", "rjrate")]
names(obs6) <- c("ID", "AGE", "COMP_Uber", "COMP_Lyft", "WEEK.TRIPS", "WEEK.TRIPS.RJ", "GENDER", "HISPANIC", "RACE", "IMMG", "EDU", "STDT", "EMPL", "HHSIZE", "CHILD", "POVERTY", "HHINCOME", "INDVINCOME", "SOURCE", "RESP", "DMS", "DMW", "RT", "LT", "PR", "MD", "rjrate")
obs6$obs <- 6

obs <- rbind(obs1, obs2, obs3, obs4, obs5, obs6)

exp_data <- obs %>% drop_na(obs, WEEK.TRIPS, WEEK.TRIPS.RJ, GENDER, EDU, RESP)

rownames(exp_data) <- 1:nrow(exp_data)

```

## Variable manipulation
```{R}
exp_data$ID <- as.factor(exp_data$ID)
exp_data$DMS <- as.numeric(exp_data$DMS)
exp_data$MD <- as.factor(exp_data$MD)
exp_data$LT <- as.factor(exp_data$LT)
exp_data$AGE <- as.factor(exp_data$AGE)
exp_data$EDU <- as.factor(exp_data$EDU)
exp_data$EMPL <- as.factor(exp_data$EMPL)
exp_data$GENDER <- as.factor(exp_data$GENDER)
exp_data$COMP_Uber <- as.factor(exp_data$COMP_Uber)
exp_data$COMP_Lyft <- as.factor(exp_data$COMP_Lyft)
exp_data$POVERTY <- as.factor(exp_data$POVERTY)

describe(exp_data)
```

## Descriptive analysis
## Socio-demographics
```{R}
WideData <- WideData %>% mutate(COMP = ifelse(COMP_Uber == 1 & COMP_Lyft == 0, 0,
                                             ifelse(COMP_Uber == 0 & COMP_Lyft == 1, 1, 2)))

GENDERcount <- WideData %>% dplyr::count(WEEK.TRIPS)

GENDERcount
```

## Description : Response to a request
```{R}
RaceCountG <- exp_data %>% dplyr::count(RESP)
RaceCountG$fraction <- RaceCountG$n/sum(RaceCountG$n)
RaceCountG

s1 <-ggplot(data=RaceCountG, aes(x=RESP, y=fraction)) + theme_bw() + 
  scale_fill_brewer(palette="Greys") + 
  geom_bar(stat="identity", fill="#4b2e83", color="#4b2e83", alpha = 0.75, width=0.3) + theme(legend.position = "bottom") +
  ylim(0,1) + 
  xlim(-0.4,1.4) + 
  ggtitle("B") + 
  xlab("Response to a request - decline vs accept") + 
  ylab("Percentage (%)") 
s1
```

## Description : rejection rate
```{R}
widedata <- exp_data

rjrateplot <- aggregate(widedata$rjrate, by=list(Category=widedata$ID), FUN=mean)

rjrateplot <- na.omit(rjrateplot)

rj0 <- filter(rjrateplot, x == 0)

nrow(rj0)/nrow(rjrateplot)

describe(rjrateplot$x)

s2 <- ggplot(data=rjrateplot, aes(x)) + 
  geom_density(fill="#4b2e83", color="#4b2e83", alpha=0.75) + 
  geom_vline(xintercept = 5.91, color = "red", alpha = 0.75) + 
  geom_text(aes(x=12, label="Mean = 5.91", y=0.075), colour="#4b2e83", alpha = 0.8) + 
  ggtitle("A") + 
  xlab("Self-reported rejection rate in past week (%)") + 
  theme_bw()

s2
```

## Multicollinearity
```{R}
cln <- lm(HHSIZE ~ DMS + DMW + RT + LT + PR + MD + POVERTY + 
             GENDER + EDU + AGE + COMP_Uber + COMP_Lyft + rjrate + EMPL, data = exp_data)
vif(cln)
```

## Logistic regression model
## Dependent variable : response to trip request
## Independent variables : Surge price (DMS), Average waiting time (DMW), Relocation time (RT), Long trip (LT), Passenger ratings (PR),
## Shared or solo trip (MD), Household size (HHSIZE), Income (POVERTY), rejection rate (rjrate), Age (AGE), Employment status (EMPL),
## Education level (EDU), Gender (GENDER), Uber (COMP_Uber), Lyft (COMP_Lyft)
## Note : to compare with other models, here we use GAM to implement logistic regression. 
```{R}
gam_y_linear <- gam(RESP ~ DMS + 
               DMW + 
               RT + 
               LT + 
               PR +
               MD + 
               HHSIZE + 
               LT*RT + 
               POVERTY + 
               rjrate  + 
               AGE + 
               EMPL + 
               EDU + 
               GENDER + 
               COMP_Uber + 
               COMP_Lyft, 
               data = exp_data, 
               family = "binomial")

summary(gam_y_linear)

AIC(gam_y_linear)

```

## Generalized additive model
## Dependent variable : response to trip request
## Independent variables : Surge price (DMS), Average waiting time (DMW), Relocation time (RT), Long trip (LT), Passenger ratings (PR),
## Shared or solo trip (MD), Household size (HHSIZE), Income (POVERTY), rejection rate (rjrate), Age (AGE), Employment status (EMPL),
## Education level (EDU), Gender (GENDER), Uber (COMP_Uber), Lyft (COMP_Lyft)
```{R}
gam_y_gam <- gam(RESP ~ 
               DMS + 
               DMW + 
               s(RT, k = 5, by = LT) + 
               LT + 
               PR +
               MD + 
               HHSIZE + 
               POVERTY + 
               s(rjrate)  + 
               AGE + 
               EMPL + 
               EDU + 
               GENDER + 
               COMP_Uber + 
               COMP_Lyft, data = exp_data, family = "binomial")

summary(gam_y_gam)

gam.check(gam_y_gam)

AIC(gam_y_gam)

BIC(gam_y_gam)

plot(gam_y_gam, pages = 1, scheme = 2, shade = TRUE)

```

## Generalized additive mixed model : used in the paper
## Dependent variable : response to trip request
## Independent variables : Surge price (DMS), Average waiting time (DMW), Relocation time (RT), Long trip (LT), Passenger ratings (PR),
## Shared or solo trip (MD), Household size (HHSIZE), Income (POVERTY), rejection rate (rjrate), Age (AGE), Employment status (EMPL),
## Education level (EDU), Gender (GENDER), Uber (COMP_Uber), Lyft (COMP_Lyft)
```{R fig, fig.height = 4.5, fig.width = 4.5}
gam_y_gamm <- gam(RESP ~ 
               DMS + 
               DMW + 
               s(RT, k = 5, by = LT) + ## smooth term one
               LT + 
               PR +
               MD + 
               HHSIZE + 
               POVERTY + 
               s(rjrate)  + ## smooth term two
               AGE + 
               EMPL + 
               EDU + 
               GENDER + 
               COMP_Uber + 
               COMP_Lyft + 
               s(ID,bs="re"), ## random effect
               data = exp_data, 
               family = "binomial")

## model results
summary(gam_y_gamm)

## variance component
gam.vcomp(gam_y_gamm, conf.lev=.95)

## model checking
gam.check(gam_y_gamm, rep = 500)

## smooth term plots
plot(gam_y_gamm, pages = 1, scheme = 2, shade = TRUE)

## final loglikelihood
logLik(gam_y_gamm)

## AIC and BIC
AIC(gam_y_gamm)
BIC(gam_y_gamm)

```

## Prediction : elasticity
```{R  fig, fig.height = 2, fig.width = 4}
library(mgcv)

exp_data_test <- exp_data
exp_data_test$RT <- exp_data$RT + 1

pred_linear <- predict.gam(gam_y_gamm, exp_data)

pred_linear_test <- predict.gam(gam_y_gamm, exp_data_test)

pred <-as.data.frame(cbind(exp_data$RT, exp_data$LT, pred_linear, pred_linear_test))

pred <- na.omit(pred)

pred$diff <- (pred$pred_linear - pred$pred_linear_test)/0.082

pred_long <- filter(pred, V2 == 2)
pred_short <- filter(pred, V2 == 1)

predplot_long <- aggregate(pred_long$diff, by=list(Category=pred_long$V1), FUN=mean)
predplot_short <- aggregate(pred_short$diff, by=list(Category=pred_short$V1), FUN=mean)

p1 <- ggplot(data = predplot_short, aes(Category, x)) + geom_point(colour="#4b2e83") + theme_bw() + ylim(-0.5, 4) + ggtitle("Short trip") + xlab("Pick-up time (min)") + ylab("WTA ($)")

p2 <- ggplot(data = predplot_long, aes(Category, x)) + geom_point(colour="#4b2e83") + theme_bw() + ylim(-0.5, 4) + ggtitle("Long trip") + xlab("Pick-up time (min)") + ylab("WTA ($)")

grid.arrange(p1, p2, nrow =1)
```

